---
title: "pums_single_year_data"
format: html
editor: visual
---

## Abstract

The American Community Survey's Public Use Microdata Sample (ACS PUMS) via the Census API releases data that we want to develop into a robust and reprodubicle pipeline for aquiring person-level data. We want to functionalize the process so we can repeat the API calls and return them in a clean and useable format. We construct a pipeline, inhest the JSON formatted data into a tabluar format suitable for downstream web based graphics in an automated fashion.

We approach this through three pillars: 1) Validation: For inputs and API components. 2) API Subsetting: For data to maximize workflows and analysis. 3) Functionalization: Around components and pipelines.

Our end goal remains a small production style foundation for nascent data exploration and analysis. APIs typically bring tremedouns functionality to organizations. However, there are many instances where too much data or functions can impact the data analysis and thus the statistical work in which we need to build.

## What are we Building

We are delivering a package style R foundation for data exploration and analysis. Our code does: 1) Writes valid API URLs to API data querying. 2) Pulls JSON payloads and converts to formatted tibbles. 3) Applies respective survey weights in plottting function. 4) Develops helper functions to assist in data formatting, aggregation and querying. 5) Scales to years and georgraphies relative to data requirements.

## Key Architecture Features

1)  URL Functions
2)  Input Helpers/Functions
3)  Plotting Functions

## Assumptions

Years Allowed: Only allows for year ranges of 2010-2022 and throws an error if not. Surveys: Single yeatr estimates preffered. Variables: Numeric variables requested. Geography: Specific code for region and time weights.

## High-Level Motivation

We built this pipeline to demonstrate how to turn the ACS PUMS endpints into a reproducible dataset suitablke for data exploration and analysis. We have enforced input validations by factoring URL and standarizing the JSON payloads. These pieces get passed into the tibble includuing factors, weights and other attributes. We have created this in a transparent manner via Guthub to show the larger thinking and development process.

Sources: Data: U.S. Census Bureau. ACS PUMS Microdata

```{r}

library(httr)
library(jsonlite)
library(tibble)
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(hms)
library(lubridate)


```

```{r}
# Build API URL with server‑side filters
.load_pums_url <- function(year = 2023,
                           numeric_vars = c("AGEP"),
                           categorical_vars = c("SEX"),
                           geography_level = c("All","Region","Division","State"),
                           geography_subset = list(),
                           survey = "acs1",
                           key = NULL) {

  geography <- match.arg(geography_level)
  subset    <- geography_subset
  key_safe  <- if (is.null(key)) "" else as.character(key)

# --- simple constants ---------------------------------------------------------
.valid_numeric     <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
.valid_categorical <- c("FER", "HHL", "HISPEED", "JWTRNS", "SCH", "SCHL", "SEX")
.valid_geos        <- c("All", "Region", "Division", "State")

.valid_region_codes   <- c(1, 2, 3, 4)
.valid_division_codes <- 1:9
.valid_state_fips     <- c(
  1,2,4,5,6,8,9,10,11,12,13,15,16,17,18,19,20,21,22,23,
  24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,
  42,44,45,46,47,48,49,50,51,53,54,55,56
)

# --- year ---------------------------------------------------------------------
.allowed_year <- function(x) {
  is.numeric(x) && length(x) == 1L && !is.na(x) && x >= 2010 && x <= 2022
}

#Added new version of function
validate_year <- function(year) {
  #creating a function to validate year
  if (!(is.numeric(year) && length(year) == 1L && !is.na(year) && year >= 2010 && year <= 2022)) {
    #this will throw an error if a year entered is not in range
    stop("year must be a single number in 2010..2022", call. = FALSE)
  }
  #outputs year in integer format
  as.integer(year)
}


# --- variables ----------------------------------------------------------------
process_numeric_vars <- function(numeric_vars = c("AGEP", "PWGTP")) {
  if (!is.character(numeric_vars)) stop("numeric_vars must be character")

  # ensure PWGTP present
  if (!"PWGTP" %in% numeric_vars) numeric_vars <- c(numeric_vars, "PWGTP")

  # check allowed
  if (!all(numeric_vars %in% .valid_numeric))
    stop("numeric_vars contains invalid names")

  # need at least one other than PWGTP
  others <- setdiff(numeric_vars, "PWGTP")
  if (length(others) == 0) stop("include at least one numeric var other than PWGTP")

  unique(numeric_vars)
}

process_categorical_vars <- function(categorical_vars = c("SEX")) {
  if (!is.character(categorical_vars)) stop("categorical_vars must be character")
  if (!all(categorical_vars %in% .valid_categorical))
    stop("categorical_vars contains invalid names")
  unique(categorical_vars)
}

# --- geography ----------------------------------------------------------------
validate_geo <- function(geography) {
  if (!is.character(geography) || length(geography) != 1L)
    stop("geography must be a single string")
  if (!(geography %in% .valid_geos))
    stop("geography must be one of All, Region, Division, State")
  geography
}

# --- subset -------------------------------------------------------------------
# subset like: list(REGION = c(1,2), DIVISION = 9, ST = c(36,42))
validate_subset <- function(subset, geography) {
  if (is.null(subset) || identical(subset, list())) return(list())

  if (!is.list(subset) || is.null(names(subset)))
    stop("subset must be a named list")

  allowed_keys <- c("REGION", "DIVISION", "ST")
  bad <- setdiff(names(subset), allowed_keys)
  if (length(bad) > 0) stop("subset has invalid names")

  # coerce to integer and dedupe
  subset <- lapply(subset, function(v) unique(as.integer(v)))

  # light checks
  if (!is.null(subset$REGION)) {
    if (any(is.na(subset$REGION)) || any(!subset$REGION %in% .valid_region_codes))
      stop("REGION must be in 1,2,3,4")
  }
  if (!is.null(subset$DIVISION)) {
    if (any(is.na(subset$DIVISION)) || any(!subset$DIVISION %in% .valid_division_codes))
      stop("DIVISION must be in 1..9")
  }
  if (!is.null(subset$ST)) {
    if (any(is.na(subset$ST)) || any(!subset$ST %in% .valid_state_fips))
      stop("ST must be valid 2-digit state FIPS")
  }

  # keep it consistent with chosen geography
  if (geography == "Region"  && any(c("DIVISION","ST") %in% names(subset)))
    stop("for Region, only use REGION")
  if (geography == "Division" && any(c("REGION","ST") %in% names(subset)))
    stop("for Division, only use DIVISION")
  if (geography == "State"   && any(c("REGION","DIVISION") %in% names(subset)))
    stop("for State, only use ST")

  subset
}


# Print everything with print()
print(paste("year =", year))
print(paste("numeric_vars =", paste(numeric_vars, collapse = ", ")))
print(paste("categorical_vars =", paste(categorical_vars, collapse = ", ")))
print(paste("geo =", geography))
print(paste("subset =", subset))
print(paste("key =", if (nzchar(key)) "Avaiable" else "Missing"))
  

# compose GET fields
get_fields <- unique(c("PWGTP", numeric_vars, categorical_vars))
params <- list(get = paste(get_fields, collapse = ","))


if (nzchar(key_safe)) params$key <- key_safe

base <- paste("https://api.census.gov/data",
              as.character(year), "acs", survey, "pums", sep = "/")

httr::modify_url(base, query = params)

}



```

```{r}
process_api_response <- function(api_response) {
  if (api_response$status_code != 200) {
    stop("Error: API request failed with status code ", api_response$status_code)
  }

  # Convert raw content to character string
  content_string <- rawToChar(api_response$content)
  
  # Parse JSON data
  parsed_data <- fromJSON(content_string)
  
  # First row contains column names, the rest is the data
  column_names <- parsed_data[1, ]
  data_rows <- parsed_data[-1, ]
  
  # Convert to tibble
  tibble_data <- as_tibble(data_rows)
  colnames(tibble_data) <- column_names
  
  #creating new variable in order to check if variables are numeric
  valid_numeric_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
   
  #Created loop to convert proper numeric variables to numeric
  for(var in colnames(tibble_data)){
    if(var %in% valid_numeric_vars){
      tibble_data[[var]] <- as.numeric(tibble_data[[var]])
    }
  }
  
  # --- categorical to factors ---------------------------------------------------
  valid_categorical_vars <- c("FER", "HHL", "HISPEED", "JWTRNS", "SCH", "SCHL", "SEX")
  for (var in colnames(tibble_data)) {
    if (var %in% valid_categorical_vars) {
      tibble_data[[var]] <- as.factor(tibble_data[[var]])
    }
  }

  # --- labeled factors ----------------------------------------------------------
  if ("SEX" %in% colnames(tibble_data)) {
    tibble_data <- tibble_data |> mutate(SEX = factor(SEX, labels = SEX_levels))
  }
  if ("SCHL" %in% colnames(tibble_data)) {
    tibble_data <- tibble_data |> mutate(SCHL = factor(SCHL, labels = SCHL_levels))
  }
  if ("SCH" %in% colnames(tibble_data)) {
    tibble_data <- tibble_data |> mutate(SCH = factor(SCH, labels = SCH_levels))
  }
  if ("JWTRNS" %in% colnames(tibble_data)) {
    tibble_data <- tibble_data |> mutate(JWTRNS = factor(JWTRNS, labels = JWTRNS_levels))
  }
  if ("HISPEED" %in% colnames(tibble_data)) {
    tibble_data <- tibble_data |> mutate(HISPEED = factor(HISPEED, labels = HISPEED_levels))
  }
  if ("HHL" %in% colnames(tibble_data)) {
    tibble_data <- tibble_data |> mutate(HHL = factor(HHL, labels = HHL_levels))
  }
  if ("FER" %in% colnames(tibble_data)) {
    tibble_data <- tibble_data |> mutate(FER = factor(FER, labels = FER_levels))
  }

  # --- JWAP / JWDP handling -----------------------------------------------------
  if("JWAP" %in% colnames(tibble_data)) {
  tibble_data <- tibble_data |>
   left_join(JWAP_intervals, join_by(JWAP)) |>
    rename(work_arrival_time = middle_time,
           arrival_meridiem = meridiem) |>
    select(-JWAP)
  } 
  
  #Pasting correct time values in for JWDP
   if("JWDP" %in% colnames(tibble_data)) {
  tibble_data <- tibble_data |>
   left_join(JWDP_intervals, join_by(JWDP)) |>
    rename(work_departure_time = middle_time2,
           departure_meridiem = meridiem2) |>
    select(-JWDP)
  } 

  return(tibble_data)
}

```

```{r}
# Function to process time variables
process_time_variable <- function(variable_name) {
 
   # Grabbing information from API
  temp <- httr::GET(paste0("https://api.census.gov/data/2022/acs/acs1/pums/variables/", variable_name, ".json"))
  temp_list <- temp$content |> rawToChar() |> jsonlite::fromJSON()
  
  # Getting just the variable names and their values
  time_values <- temp_list$values$item
  time_vals_sorted <- time_values[sort(names(time_values))]
  
  # Converting the values into a tibble
  time_tibble <- tibble(id = seq(0, length(time_vals_sorted) - 1), value = time_vals_sorted)
  
  # Converting to wide format
  time_wide <- separate_wider_delim(time_tibble, 
                                     cols = c("value"), 
                                     delim = " ", 
                                     names = c("start", "meridiem", "to", "end", "meridiem_2"), 
                                     too_few = "debug", 
                                     names_repair = "unique", 
                                     too_many = "drop") 
  
  # Select the needed rows and convert to usable format
  time_intervals <- time_wide |>
    select(id, start, meridiem, end, meridiem_2) |> #meridiem is to track if values were for am or pm
    mutate(start = parse_hm(start),
           end = parse_hm(end),
           variable_name = id)  
  
  # Make row 1 display NA across all columns for row 1    
  time_intervals[1, ] <- NA
  
  # Creating middle time value variable
  time_intervals <- time_intervals |>
    mutate(interval = (end - start) / 2,
           interval = as.numeric(interval),
           interval = as.period(interval))

  # Turning start into a period   
  time_intervals <- time_intervals |>
    mutate(start = as.period(start),
           middle_time = as.duration(start + interval)) |>
    select(middle_time, variable_name, meridiem) 
  
  # Turning the first row variable back to 0
  time_intervals[1, 2] <- 0
          
  # Converting back to a time variable        
  time_intervals <- time_intervals |>
    mutate(middle_time = hms::hms(seconds = middle_time))
  
  return(time_intervals)
} 

#Renaming variables to insure they work in the helper function
JWAP_intervals <- process_time_variable("JWAP")
JWAP_intervals <- JWAP_intervals |>
  rename(JWAP = variable_name)
         
JWDP_intervals <- process_time_variable("JWDP")
JWDP_intervals <- JWDP_intervals |>
  rename(JWDP = variable_name,
         middle_time2 = middle_time,
         meridiem2 = meridiem)

```

```{r}
# Function to process categorical variables
process_cat_variable <- function(variable_name) {
 
   # Grabbing information from API
  temp <- httr::GET(paste0("https://api.census.gov/data/2022/acs/acs1/pums/variables/", variable_name, ".json"))
  temp_list <- temp$content |> rawToChar() |> jsonlite::fromJSON()
  
  # Getting just the variable names and their values
  cat_values <- temp_list$values$item
  cat_vals_sorted <- cat_values[sort(names(cat_values))]
  
  # Converting the values into a tibble
  cat_tibble <- tibble(id = seq(0, length(cat_vals_sorted) - 1), value = cat_vals_sorted)
  cat_vector <- cat_tibble$value
  
  return(cat_vector)
}

#Running function for categorical variables to store levels
SCHL_levels <- process_cat_variable("SCHL")
SEX_levels <- process_cat_variable("SEX")
SCH_levels <- process_cat_variable("SCH")
JWTRNS_levels <- process_cat_variable("JWTRNS")
HISPEED_levels <- process_cat_variable("HISPEED")
HHL_levels <- process_cat_variable("HHL")
FER_levels <- process_cat_variable("FER")
```

```{r}
# Add comments
.get_pums_tibble <- function(url, show_call = FALSE) {
  if (isTRUE(show_call)) print(paste("PUMS URL:", url))

  resp <- httr::GET(url)
  httr::stop_for_status(resp)

  parsed <- jsonlite::fromJSON(rawToChar(resp$content))
  hdr <- parsed[1, ]
  df  <- as.data.frame(parsed[-1, ], stringsAsFactors = FALSE, check.names = FALSE)
  names(df) <- hdr
  tibble::as_tibble(df, .name_repair = "minimal")
}


```

```{r}

```

```{r}
year <- 2022
numeric_vars <- c("AGEP", "JWMNP")
categorical_vars <- c("SEX")
geography <- "State"            
subset <- list(ST = c(37, 36))  
key <- "2c10d8b7f4f3ca11ddef8ced42f41817fa34d9f7"

url <- .load_pums_url(
  year = year,
  numeric_vars = numeric_vars,
  categorical_vars = categorical_vars,
  geography_level = geography,
  geography_subset = subset,
  survey = "acs1",
  key = key
)

tbl <- .get_pums_tibble(url, show_call = TRUE)

print(tbl)


```

```{r}
# Function for multiple years
query_multiple_years <- function(
    years,
    numeric_vars = c("AGEP", "PWGTP"),
    categorical_vars = c("SEX"),
    geography_level = "All",
    geography_subset = NULL,
    survey = "acs1",
    key = NULL)
  {
  all_years_data <- list()
 
  for (yr in years) {
    # validate year inside loop
    if (yr < 2010 || yr > 2022) {
      warning(paste("Skipping invalid year:", yr))
      next
    }
   
    # Build URL for single year
    url <- .load_pums_url(
      year = yr,
      numeric_vars = numeric_vars,
      categorical_vars = categorical_vars,
      geography_level = geography_level,
      geography_subset = geography_subset,
      survey = survey,
      key = key
    )
   
    # Fetch raw API response
    resp <- httr::GET(url)
    httr::stop_for_status(resp)
   
    # Process into cleaned tibble
    tbl <- process_api_response(resp)
   
    # Add year column so you know which year rows came from
    tbl$year <- yr
   
    # Save in list
    all_years_data[[as.character(yr)]] <- tbl
  }
 
  # Combine into single tibble
  final_data <- bind_rows(all_years_data)
  class(final_data) <- c("census", class(final_data))
 
  return(final_data)
}

```

```{r}
#year <- c( 2016, 2017, 2018)
year <- c( 2016, 2017)
numeric_vars <- c("AGEP", "PWGTP", "JWAP", "GRPIP" )
categorical_vars <- c("SEX", "HISPEED", "SCH", "FER")
geography <- "State"            
subset <- list(ST = c(50, 56))  
key <- "2c10d8b7f4f3ca11ddef8ced42f41817fa34d9f7"

multi_year_result <- query_multiple_years(
  year = year,
  numeric_vars = numeric_vars,
  categorical_vars = categorical_vars,
  geography_level = geography,
  geography_subset = subset,
  survey = "acs1",
  key = key
)

tbl <- .get_pums_tibble(url, show_call = TRUE)

print(tbl)

```

```{r}
# Define custom summary function for the 'census' class
summary.census <- function(census_data, 
                           numeric_vars = NULL, 
                           categorical_vars = NULL) {
  
  # Ensure PWGTP exists and is numeric
  if (!"PWGTP" %in% names(census_data)) {
    stop("Weight variable 'PWGTP' is missing from the dataset.")
  }
  
  # Separate numeric and categorical columns from the data
  if (is.null(numeric_vars)) {
    numeric_vars <- names(census_data)[sapply(census_data, is.numeric) & names(census_data) != "PWGTP"]
  }
  
  if (is.null(categorical_vars)) {
    categorical_vars <- names(census_data)[sapply(census_data, is.factor)]
  }
  # Weighted mean and standard deviation calculations
  summarize_numeric <- function(var_name, data) {
    numeric_vector <- as.numeric(data[[var_name]])
    weight_vector <- as.numeric(data[["PWGTP"]])
    
    # Check for missing values in the weight vector
    if (any(is.na(weight_vector))) {
      stop("Missing values found in the weight variable 'PWGTP'.")
    }
    
    # Calculate weighted mean
    weighted_mean <- sum(numeric_vector * weight_vector, na.rm = TRUE) / sum(weight_vector, na.rm = TRUE)
    
    # Calculate weighted standard deviation
    weighted_var <- sum(numeric_vector^2 * weight_vector, na.rm = TRUE) / sum(weight_vector, na.rm = TRUE)
    weighted_sd <- sqrt(weighted_var - weighted_mean^2)
    
    return(list(mean = weighted_mean, sd = weighted_sd))
  }
  
  # Initialize a list to store the summary results
  summary_list <- list()
  
  # Summarize numeric variables
  for (var in numeric_vars) {
    summary_list[[var]] <- summarize_numeric(var, census_data)
  }
  
  # Summarize categorical variables (counts)
  for (var in categorical_vars) {
    summary_list[[var]] <- table(census_data[[var]], useNA = "ifany")
  }
  
  return(summary_list)
}
```

```{r}
#Creating function for user to create weighted box plots
plot.census <- function(tibble_data, categorical_var, numeric_var) {
  
  #Telling user they must specify at least one categorical variable
  if (missing(categorical_var) || missing(numeric_var)){
    stop("Must include one categorical variable and one numeric variable")
  }
  #Telling user time variables can not be use as numeric variables
  if ("JWAP" %in% numeric_var){
    stop("JWAP is a time variable")
  }
   if ("JWDP" %in% numeric_var){
    stop("JWDP is a time variable")
  }
  
  p <- ggplot(tibble_data, 
              aes(x = get(categorical_var), y = get(numeric_var), weight = PWGTP)) + geom_boxplot() + 
                labs(title = paste("Box plot of", deparse(substitute(categorical_var)), "by", deparse(substitute(numeric_var))), x = categorical_var, y = numeric_var) + theme_light() 
  
  print(p)
}

# Take a reproducible sample of 10,000 rows
set.seed(123)   # ensures reproducibility
multi_year_sample <- dplyr::sample_n(multi_year_result, 100000)

# Now plot using the sample
plot.census(multi_year_sample, "SEX", "AGEP")
#plot.census(multi_year_result, "SEX", "AGEP")
```
